---
title: "Lab09Report"
author: "Lyssa Erskine"
format: pdf
editor: visual
---

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)
library(corrplot)
library(plotly)
library(skimr)

candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```
**Q1. How many different candy types are in this dataset?**

There are 85 candy types

```{r}
nrow(candy)
```
**Q2. How many fruity candy types are in the dataset?**

There are 38 fruity candies
```{r}
sum(candy$fruity)
```
**Q3. What is your favorite candy (other than Twix) and its winpercent?**

My favorite candy is Snickers and its winpercent is 76.67

```{r}
candy["Snickers", "winpercent"]
```
**Q4. What is the winpercent for Kit Kat?**

The winpercent is 76.77 for kit kat. 
```{r}
candy["Kit Kat", "winpercent"]
```
**Q5. What is the winpercent for Tootsie Roll Snack Bars?**

The winpercent is 49.65 for Tootsie Rolls
```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```
**Q6. Is any variable on a different scale?**

Yes. winpercent is on a different scale (0–100), while most other variables are binary (0 or 1).

**Q7. What do 0 and 1 represent in chocolate?**

1 = contains chocolate
0 = does not contain chocolate

**Q8. Plot a histogram of winpercent**
```{r}
hist(candy$winpercent)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins = 10)
```
**Q9. Is the distribution symmetrical?**
No, it is slightly skewed to the left. 

**Q10. Is the center above or below 50%?**
Slightly below 50%.

**Q11. Is chocolate or fruity candy ranked higher?**
Chocolate candy is ranked higher on average.
```{r}
mean(candy$winpercent[candy$chocolate == 1])
mean(candy$winpercent[candy$fruity == 1])
```
**Q12. Is this difference statistically significant?**
Yes, it is statistically significant.
```{r}
t.test(candy$winpercent[candy$chocolate == 1],
       candy$winpercent[candy$fruity == 1])
```
**Q13. What are the five least liked candies?**
Nik L Nip
Boston Baked Beans
Chiclets
Super Bubble
Jawbusters

```{r}
head(candy[order(candy$winpercent), ], 5)
```
**Q14. What are the top 5 favorite candies?**
Reese’s Peanut Butter Cup
Reese’s Miniatures
Twix
Kit Kat
Snickers
```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], 5)
```
**Q15. Make a barplot of rankings**
```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
**Q16. Improve with sorting**
```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
**Q17. Worst ranked chocolate candy?**
Sixlets

**Q18. Best ranked fruity candy?**
Starburst

**Q19. Best candy for lowest price (“bang for buck”)?**
Reese’s Miniatures

**Q20. Top 5 most expensive candies & least popular among them**
Most expensive:
Nik L Nip
Nestle Smarties
Ring Pop
Hershey’s Krackel
Hershey’s Milk Chocolate

Least popular: Nik L Nip

```{r}
head(candy[order(candy$pricepercent, decreasing = TRUE), ], 5)
```
**Q21. Lollipop chart**
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent),
                   xend = 0)) +
  geom_point()
```
**Q22. Which variables are anti-correlated?**
chocolate and fruity

**Q23. Which variables contribute most to PC1?**
Chocolate (negative direction)
Fruity (positive direction)
Bar
Winpercent

**Q24. PCA loadings plot**
```{r}
pca <- prcomp(candy, scale = TRUE)

ggplot(as.data.frame(pca$rotation)) +
  aes(x = PC1, y = reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```
Positive PC1: fruity, hard candy
Negative PC1: chocolate, bar, winpercent

**Q25. What makes a “winning” candy?**
A winning candy is typically:

Chocolate-based
Often a candy bar
Not fruity or hard candy

Bar plots show top candies are chocolate (Reese’s, Twix, Kit Kat)
Correlation shows chocolate positively linked with popularity
PCA shows chocolate vs fruity separation as main pattern


